---
title: "Suspensions"
editor_options: 
  chunk_output_type: console
---

## A Track of Players and Their Performances With Suspensions (Doping)

Our analysis focuses on the impact of doping sanctions on player performance in professional tennis. We've collected data from both the International Tennis Integrity Agency (ITIA) and Wikipedia to create a comprehensive dataset of doping cases.


## Methods

We faced several challenges in data collection:
* Lack of a centralized database for tennis doping cases
* Inconsistent reporting of suspension dates and details across sources
* Difficulty in matching player names between doping data and match results

We scraped data from two primary sources:
1. The International Tennis Integrity Agency (ITIA) website, using web scraping techniques.
2. Wikipedia's list of doping cases in tennis.

Then, we merged the doping dataset with our existing ATP match data. This process required careful cleaning and standardization of player names and dates.

```{r lib, echo = FALSE, warning = FALSE, message = FALSE}

library(tidyverse)
library(rvest)
library(httr)
library(dplyr)
library(stringr)
library(ggplot2)
library(maps)
library(plotly)
library(leaflet)
library(sf)

#install.packages("rnaturalearthdata")

```



```{r setup, echo = FALSE, warning = FALSE, message = FALSE}

# Scrape data from ITIA sanctions page
url <- "https://www.itia.tennis/sanctions/"
page <- read_html(url)

# Extract and process sanction data
itia_sanctions <- page %>%
  html_nodes(".accordion__item") %>%
  map_df(function(item) {
    # Extract player name
    name_element <- item %>% 
      html_node(".accordion__column:nth-child(1)")
    
    surname <- name_element %>%
      html_node(".sort1") %>%
      html_text(trim = TRUE)
    
    full_name <- name_element %>%
      html_text(trim = TRUE)
    player_name <- str_trim(str_replace(full_name, paste0("^", surname), ""))
    
    # Extract program, sanction, and dates
    program <- item %>% 
      html_node(".accordion__column:nth-child(2)") %>% 
      html_text(trim = TRUE)
    
    sanction <- item %>% 
      html_node(".accordion__column:nth-child(3)") %>% 
      html_text(trim = TRUE)
    
    dates <- item %>% 
      html_node(".accordion__item-content") %>% 
      html_text(trim = TRUE) %>%
      str_extract_all("\\d{2}/\\d{2}/\\d{4}") %>%
      unlist()
    
    start_date <- if(length(dates) >= 1) dates[1] else NA
    end_date <- if(length(dates) >= 2) dates[2] else NA
    
    tibble(player_name = player_name, 
      program = program, 
      sanction = sanction, 
      start_date = start_date,
      end_date = end_date
    )
  })


# Function to convert words to numbers
words_to_numbers <- function(word) {
  word_map <- c("one" = 1, "two" = 2, "three" = 3, "four" = 4, "five" = 5,
                "six" = 6, "seven" = 7, "eight" = 8, "nine" = 9, "ten" = 10,
                "eleven" = 11, "twelve" = 12, "thirteen" = 13, "fourteen" = 14,
                "fifteen" = 15, "sixteen" = 16, "seventeen" = 17, "eighteen" = 18,
                "nineteen" = 19, "twenty" = 20, "thirty" = 30, "forty" = 40,
                "fifty" = 50, "sixty" = 60, "seventy" = 70, "eighty" = 80,
                "ninety" = 90)
  
  return(as.character(word_map[tolower(word)]))
}


# Clean and process ITIA sanctions data
itia_sanctions <- itia_sanctions %>%
  # Only include TADP sanctions
  filter(program == "TADP") %>%
  
  # Convert dates to proper Date objects
  mutate(start_date = as.Date(start_date, format = "%d/%m/%Y"),
         end_date = as.Date(end_date, format = "%d/%m/%Y")) %>%
  
  # Create a separate sanction_type column
  mutate(sanction_type = case_when(
    str_detect(sanction, "ban") ~ "Ban",
    str_detect(sanction, "suspension") ~ "Suspension",
    TRUE ~ "Other"
  )) %>%
  
  # Create timeframe column
  mutate(time_frame = case_when(
      str_detect(sanction, "Lifetime") ~ "Lifetime",
      str_detect(sanction, "Provisional") ~ "Provisional",
      TRUE ~ str_extract(sanction, "(\\d+|\\w+)\\s+(years?|months?)(\\s+and\\s+(\\d+|\\w+)\\s+(years?|months?))?")
    ),
    time_frame = str_replace_all(time_frame, "\\b(\\w+)\\b", function(x) {
      num <- words_to_numbers(x)
      if (!is.na(num)) num else x
    })
  )

print(head(itia_sanctions))


# Scrape doping cases from Wikipedia
url <- "https://en.wikipedia.org/wiki/Category:Doping_cases_in_tennis"
page <- read_html(url)

# Extract the list of players
players <- page %>%
  html_nodes(".mw-category-group li a") %>%
  html_text()

wiki_doping<- data.frame(player_name = players)
head(wiki_doping)


# Merge ITIA and Wikipedia doping data
doping_data <- full_join(wiki_doping, itia_sanctions) %>%
  distinct() %>%
  mutate(sanction_start = as.Date(start_date),
    sanction_end = as.Date(end_date))


# Read tennis match data
tennis_data <- read_csv("data/cleaned_df_webscraping_data_added.csv")

# Extract unique player names from the cleaned tennis data
tennis_players <- unique(c(tennis_data$winner_name, tennis_data$loser_name))

# Filter doping data to include only players in tennis data
filtered_doping_data <- doping_data %>%
  filter(player_name %in% tennis_players)

# Merge doping data with tennis match data
merged_data <- tennis_data %>%
  left_join(doping_data, by = c("winner_name" = "player_name")) %>%
  rename(winner_doping_start = start_date, 
         winner_doping_end = end_date,
         winner_sanction = sanction) %>%
  left_join(doping_data, by = c("loser_name" = "player_name")) %>%
  rename(loser_doping_start = start_date,
         loser_doping_end = end_date,
         loser_sanction = sanction)

# Remove rows where there's no doping data for either player
final_data <- merged_data %>%
  filter(!is.na(winner_doping_start) | !is.na(loser_doping_start))


```



```{r playerstats, echo = FALSE, warning = FALSE, message = FALSE}

## Player performance statistics
# Filter for matches involving players with doping records
doping_matches <- final_data %>%
  filter(winner_name %in% doping_data$player_name | loser_name %in% doping_data$player_name)

# Calculate player performance statistics
all_player_performance <- final_data %>%
  pivot_longer(cols = c(winner_name, loser_name), names_to = "role", values_to = "player_name") %>%
  group_by(player_name) %>%
  reframe(matches_total = n(),
    wins = sum(role == "winner_name"),
    losses = sum(role == "loser_name"),
    win_rate = wins / matches_total,
    in_doping_dataset = player_name %in% doping_data$player_name) %>%
  distinct()


```


## Timeline of Doping Cases

```{r timeline, echo = FALSE, warning = FALSE, message = FALSE}

## Timeline Data
# Prepare timeline data for doping cases
timeline_data <- doping_data %>%
  filter(!is.na(start_date)) %>%
  arrange(start_date)

# Plot timeline of doping cases
timeline_plot <- ggplot(timeline_data, aes(x = start_date, y = reorder(player_name, start_date))) +
  geom_point(color = "#AF125A", size = 3) +
  geom_segment(aes(xend = end_date, yend = player_name), color = "#F390BE", size = 1) +
  labs(title = "Timeline of Doping Cases in Tennis",
    x = "Date",
    y = "Player Name") +
  theme_minimal(base_size = 12) + 
  aes(text = paste("Player:", player_name, "<br>Sanction Start Date:", start_date, "<br>Sanction End Date:", end_date))

ggplotly(timeline_plot, tooltip = "text")


```


## Sanction Durations

```{r sanctionduration, echo = FALSE, warning = FALSE, message = FALSE}

## Sanction Durations
# Analyze sanction durations
sanction_analysis <- doping_data %>%
  mutate(sanction_duration_days = as.numeric(end_date - start_date)) %>%
  filter(!is.na(sanction_duration_days))

# Visualize sanction durations
sanction_plot <- ggplot(sanction_analysis, aes(x = reorder(player_name, -sanction_duration_days), y = sanction_duration_days)) +
  geom_bar(stat = "identity", fill = "#AF125A") +
  labs(title = "Sanction Durations for Doping Cases",
    x = "Player Name",
    y = "Duration (Days)") +
  theme_classic(base_size = 12) +
  coord_flip() +
  aes(text = paste("Player:", player_name, "<br>Sanction Duration:", sanction_duration_days, "days"))

ggplotly(sanction_plot, tooltip = "text")


```


## Case Study: Mikael Ymer

```{r ymer, echo = FALSE, warning = FALSE, message = FALSE}

## Analyze Mikael Ymer's performance
# Look at Mikael Ymer's performance stats 
pre_sanction_performance <- final_data %>%
  filter((winner_name == "Mikael Ymer" | loser_name == "Mikael Ymer") & 
         tourney_date < as.Date("2023-07-18")) %>%
  summarize(matches_played = n(),
    wins = sum(winner_name == "Mikael Ymer"),
    win_rate = wins / matches_played)

ymer_career_performance <- final_data %>%
  filter(winner_name == "Mikael Ymer" | loser_name == "Mikael Ymer") %>%
  summarize(matches_played = n(),
    wins = sum(winner_name == "Mikael Ymer"),
    win_rate = wins / matches_played)

ymer_performance_over_time <- final_data %>%
  filter(winner_name == "Mikael Ymer" | loser_name == "Mikael Ymer") %>%
  mutate(won = winner_name == "Mikael Ymer") %>%
  arrange(tourney_date) %>%
  mutate(cumulative_win_rate = cummean(won))

# Plot Mikael Ymer's performance over time
ggplot(ymer_performance_over_time, aes(x = tourney_date, y = cumulative_win_rate)) +
  geom_line(color = "blue", size = 0.4) +
  geom_vline(xintercept = as.Date("2023-07-18"), color = "#AF125A", linetype = "dashed") +
  labs(title = "Mikael Ymer's Cumulative Win Rate Over Time",
    x = "Date",
    y = "Cumulative Win Rate",
    caption = "Red dashed line indicates start of sanction") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1))

```

In the case of Mikael Ymer, we noticed a slight decline in performance leading up to his sanction, which could be an interesting area for further study. 


## Mapping Doping Cases by Country

```{r mapdoping, echo = FALSE, warning = FALSE, message = FALSE}

## Mapping Doping Cases
# Prepare data for geographic distribution of doping cases
ioc_country_mapping <- c(
  "USA" = "United States",
  "GBR" = "United Kingdom",
  "FRA" = "France",
  "ESP" = "Spain",
  "ITA" = "Italy",
  "GER" = "Germany",
  "AUS" = "Australia",
  "JPN" = "Japan",
  "CHN" = "China",
  "RUS" = "Russia",
  "SRB" = "Republic of Serbia",
  "SWE" = "Sweden",
  "SVK" = "Slovakia")

country_data <- merged_data %>%
  filter(winner_name %in% doping_data$player_name | loser_name %in% doping_data$player_name) %>%
  mutate(winner_doping = winner_name %in% doping_data$player_name,
    loser_doping = loser_name %in% doping_data$player_name) %>%
  pivot_longer(cols = c(winner_ioc, loser_ioc),
    names_to = "player_type",
    values_to = "ioc") %>%
  filter((player_type == "winner_ioc" & winner_doping) | (player_type == "loser_ioc" & loser_doping)) %>%
  mutate(country = ioc_country_mapping[ioc]) %>%
  group_by(country) %>%
  summarize(cases_count = n_distinct(case_when(
    player_type == "winner_ioc" ~ winner_name,
    player_type == "loser_ioc" ~ loser_name
  )))

# Join with world map data
world_map <- st_as_sf(rnaturalearth::ne_countries(scale = "medium", 
  type = "countries",
  continent = NULL,
  country = NULL,
  geounit = NULL,
  sovereignty = NULL,
  returnclass = "sf"))
map_data <- world_map %>%
  left_join(country_data, by = c("subunit" = "country"))

pal <- colorNumeric(palette = "PRGn", domain = map_data$cases_count, na.color = "transparent")

leaflet(map_data) %>%
  addTiles() %>%
  addPolygons(fillColor = ~pal(cases_count),
    weight = 0.5,
    color = "black",
    opacity = 0.4,
    fillOpacity = 0.7,
    label = ~paste(name, ": ", cases_count, " cases"),
    labelOptions = labelOptions(
      style = list("font-size" = "12px", "font-weight" = "bold"),
      direction = "auto")) %>%
  addLegend(position = "bottomright",
    pal = pal,
    values = map_data$cases_count,
    title = "Doping Cases",
    opacity = 1)

```

We observed a higher concentration of doping cases in Italy and the United States, which may warrant further investigation into training practices or anti-doping measures in these countries. However, our sample was limited with only 8 countries. 


## Conclusion

Doping in professional sports remains a pervasive issue, with significant implications for player performance and career trajectory. Our analysis of doping cases in professional tennis reveals a complex landscape of sanctions, performance impacts, and geographic patterns. While the data is limited, it suggests that doping sanctions can have significant effects on players' careers and the sport as a whole. Further research is needed to fully understand the long-term implications of doping in tennis, including a profile of the specific substances, the fine amount, etc. 


