---
title: "Wins"
output: 
  html_document:
    toc: true
    toc_float: true
    css: styles.css
---

## Country/Player winning the most
```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)

tennis_data <- read.csv("cleaned_data_with_full_2024_data_added.csv")

top_countries <- tennis_data %>%
  filter(!is.na(winner_ioc) & winner_ioc != "NA") %>% # Exclude missing or "NA" countries
  count(winner_ioc) %>%
  top_n(7, n) %>%
  pull(winner_ioc)

filtered_data <- tennis_data %>%
  filter(winner_ioc %in% top_countries)

ggplot(filtered_data, aes(x = winner_ioc, fill = tourney_name)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Top 7 Countries by Wins Across Major Tournaments",
    x = "Country",
    y = "Number of Wins",
    fill = "Tournament"
  ) +
  theme_minimal()
```

We can see that the French Open was won by players from `r filtered_data %>% filter(tourney_name == 'Roland Garros') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, the Australian Open was also won by `r filtered_data %>% filter(tourney_name == 'Australian Open') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, as well as the US Open which was also won by players representing `r filtered_data %>% filter(tourney_name == 'US Open') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, finally Wimbledon was won the most often by players from the `r filtered_data %>% filter(tourney_name == 'Wimbledon') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)`. This is due to `r filtered_data %>% count(winner_name, winner_ioc, sort = TRUE) %>% slice(2) %>% summarise(names = paste(winner_name, '(', winner_ioc, ')', collapse = ', ')) %>% pull(names)` having won the second most tournaments in the past 10 years after Novak Djokovic, in addition to other spanish players having won several titles.

## Surface Matters
The Grand Slam tennis tournaments are played on different surfaces, potentially affecting player performance. The Australian and US Opens use hard courts, the French Open is played on clay, and Wimbledon is played on grass.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load necessary libraries
library(dplyr)
library(ggplot2)

# Map tournament names to surfaces and adjust counts for Hard Courts
tennis_data <- tennis_data %>%
  mutate(surface = case_when(
    tourney_name %in% c("Australian Open", "US Open") ~ "Hard Courts",
    tourney_name == "Roland Garros" ~ "Clay",
    tourney_name == "Wimbledon" ~ "Grass",
    TRUE ~ "Other"
  )) %>%
  mutate(weighted_wins = ifelse(surface == "Hard Courts", 0.5, 1))

# Count wins by country and surface, adjusting for weighted wins
top_countries <- tennis_data %>%
  group_by(winner_ioc, surface) %>%
  summarize(wins = sum(weighted_wins), .groups = "drop") %>%
  arrange(surface, desc(wins)) %>%
  group_by(surface) %>%
  slice_max(order_by = wins, n = 5)

# Create the plot
ggplot(top_countries, aes(x = reorder(winner_ioc, -wins), y = wins, fill = winner_ioc)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~surface, scales = "free_x") +
  labs(
    title = "Top 5 Countries by Wins on Each Surface",
    x = "Country",
    y = "Number of Wins",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Find the top country for each surface
top_country_summary <- top_countries %>%
  group_by(surface) %>%
  slice_max(order_by = wins, n = 1)

# Construct sentences
hard_courts_winners <- paste(top_country_summary %>% filter(surface == "Hard Courts") %>% pull(winner_ioc))
clay_courts_winners <- paste(top_country_summary %>% filter(surface == "Clay") %>% pull(winner_ioc))
grass_courts_winners <- paste(top_country_summary %>% filter(surface == "Grass") %>% pull(winner_ioc))


```

Clay court tournaments,often associated with extensive training on slower surfaces like in Spain, were most often won by players from `r top_country_summary %>% filter(surface == "Clay") %>% pull(winner_ioc)`.  
Grass court tournaments were most often won by players from `r top_country_summary %>% filter(surface == "Grass") %>% pull(winner_ioc)`.
Hard court tournaments were most often won by players from `r top_country_summary %>% filter(surface == "Hard Courts") %>% pull(winner_ioc)`.  


## Trends of Winners Over Time

Over the years, the landscape of tennis has seen significant changes in terms of who wins and how often. This first graph shows the trends in the number of titles won by players in Grand Slam tournaments, illustrating the dominance of certain players over time. As we move forward, we’ll take a closer look at how these trends correlate with players' ages, match lengths, and other factors.

```{r winners, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)

tennis_data <- read.csv("data/cleaned_df_webscraping_data_added.csv") %>%
  mutate(
    tourney_date = as.Date(tourney_date),
    winner_age = as.numeric(winner_age),
    loser_age = as.numeric(loser_age),
    minutes = as.numeric(minutes)
  )

top_winners_per_year <- tennis_data %>%
  filter(tourney_name %in% c("Australian Open", "Roland Garros", "Wimbledon", "US Open")) %>%
  drop_na(winner_name) %>%
  count(year = format(tourney_date, "%Y"), winner_name, name = "wins") %>%
  group_by(year) %>%
  slice_max(wins, n = 1) %>%
  ungroup()

ggplot(top_winners_per_year, aes(x = as.integer(year), y = wins, color = winner_name)) +
  geom_line() +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Trends of ATP Winners Over Time in Grand Slams",
    x = "Year", y = "Number of Wins"
  ) +
  scale_x_continuous(breaks = unique(as.integer(top_winners_per_year$year))) +
  theme(legend.position = "right")
```

- Novak Djokovic (green line) and Rafael Nadal (blue line) are the most prominent players, with a consistent high number of Grand Slam wins over the years.  
- Dominic Thiem achieved a win in 2020 but hasn’t sustained dominance compared to Djokovic and Nadal.  
- Jannik Sinner appears as a new entrant, indicating the rise of younger talent breaking into Grand Slam victories in recent years.  






