---
title: "Tournament Analyses"
output: 
  html_document:
    toc: true
    toc_float: true
    css: styles.css
---

## Country winning the most
```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)

tennis_data <- read.csv("data/cleaned_data_with_full_2024_data_added.csv")

top_countries <- tennis_data %>%
  filter(!is.na(winner_ioc) & winner_ioc != "NA") %>%
  count(winner_ioc) %>%
  top_n(7, n) %>%
  pull(winner_ioc)

filtered_data <- tennis_data %>%
  filter(winner_ioc %in% top_countries)

ggplot(filtered_data, aes(x = winner_ioc, fill = tourney_name)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Top 7 Countries by Wins Across Major Tournaments",
    x = "Country",
    y = "Number of Wins",
    fill = "Tournament"
  ) +
  theme_minimal()
```

We can see that the French Open was won by players from `r filtered_data %>% filter(tourney_name == 'Roland Garros') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, the Australian Open was also won by `r filtered_data %>% filter(tourney_name == 'Australian Open') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, as well as the US Open which was also won by players representing `r filtered_data %>% filter(tourney_name == 'US Open') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)` the most often, finally Wimbledon was won the most often by players from the `r filtered_data %>% filter(tourney_name == 'Wimbledon') %>% count(winner_ioc, sort = TRUE) %>% slice(1) %>% pull(winner_ioc)`. This is due to `r filtered_data %>% count(winner_name, winner_ioc, sort = TRUE) %>% slice(2) %>% summarise(names = paste(winner_name, '(', winner_ioc, ')', collapse = ', ')) %>% pull(names)` having won the second most tournaments in the past 10 years after Novak Djokovic, in addition to other spanish players having won several titles.

## Surface Matters
The Grand Slam tennis tournaments are played on different surfaces, potentially affecting player performance. The Australian and US Opens use hard courts, the French Open is played on clay, and Wimbledon is played on grass.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)

tennis_data <- tennis_data %>%
  mutate(surface = case_when(
    tourney_name %in% c("Australian Open", "US Open") ~ "Hard Courts",
    tourney_name == "Roland Garros" ~ "Clay",
    tourney_name == "Wimbledon" ~ "Grass",
    TRUE ~ "Other"
  )) %>%
  mutate(weighted_wins = ifelse(surface == "Hard Courts", 0.5, 1))

top_countries <- tennis_data %>%
  group_by(winner_ioc, surface) %>%
  summarize(wins = sum(weighted_wins), .groups = "drop") %>%
  arrange(surface, desc(wins)) %>%
  group_by(surface) %>%
  slice_max(order_by = wins, n = 5)

ggplot(top_countries, aes(x = reorder(winner_ioc, -wins), y = wins, fill = winner_ioc)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~surface, scales = "free_x") +
  labs(
    title = "Top 5 Countries by Wins on Each Surface",
    x = "Country",
    y = "Number of Wins",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
top_country_summary <- top_countries %>%
  group_by(surface) %>%
  slice_max(order_by = wins, n = 1)

hard_courts_winners <- paste(top_country_summary %>% filter(surface == "Hard Courts") %>% pull(winner_ioc))
clay_courts_winners <- paste(top_country_summary %>% filter(surface == "Clay") %>% pull(winner_ioc))
grass_courts_winners <- paste(top_country_summary %>% filter(surface == "Grass") %>% pull(winner_ioc))


```

Clay court tournaments, often associated with extensive training on slower surfaces like in Spain, were most often won by players from `r top_country_summary %>% filter(surface == "Clay") %>% pull(winner_ioc)`.  
Grass court tournaments were most often won by players from `r top_country_summary %>% filter(surface == "Grass") %>% pull(winner_ioc)`.
Hard court tournaments were most often won by players from `r top_country_summary %>% filter(surface == "Hard Courts") %>% pull(winner_ioc)`.  


## Top players for each tournament
```{r top-players, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)

top_players <- tennis_data %>%
  filter(tourney_name %in% c("Australian Open", "Roland Garros", "Wimbledon", "US Open")) %>%
  group_by(tourney_name, winner_name) %>%
  summarize(wins = n(), .groups = "drop") %>%
  arrange(tourney_name, desc(wins)) %>%
  group_by(tourney_name) %>%
  slice_max(order_by = wins, n = 5)

# Create the plot
ggplot(top_players, aes(x = reorder(winner_name, -wins), y = wins, fill = winner_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~tourney_name, scales = "free_x") +
  labs(
    title = "Top 5 Players by Wins for Each Grand Slam Tournament",
    x = "Player",
    y = "Number of Wins",
    fill = "Player"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The player with the most wins at the Australian Open is `r top_players %>% filter(tourney_name == "Australian Open") %>% slice(1) %>% pull(winner_name)` with `r top_players %>% filter(tourney_name == "Australian Open") %>% slice(1) %>% pull(wins)` wins, at Roland Garros, the top player is also `r top_players %>% filter(tourney_name == "Roland Garros") %>% slice(1) %>% pull(winner_name)` with `r top_players %>% filter(tourney_name == "Roland Garros") %>% slice(1) %>% pull(wins)` wins, at Wimbledon the top player is also `r top_players %>% filter(tourney_name == "Wimbledon") %>% slice(1) %>% pull(winner_name)` with `r top_players %>% filter(tourney_name == "Wimbledon") %>% slice(1) %>% pull(wins)` wins and at the US Open the top player is also `r top_players %>% filter(tourney_name == "US Open") %>% slice(1) %>% pull(winner_name)` with `r top_players %>% filter(tourney_name == "US Open") %>% slice(1) %>% pull(wins)` wins.

## Participation trends over time

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(readr)

participants <- tennis_data %>%
  filter(tourney_name %in% c("Wimbledon", "Roland Garros", "US Open", "Australian Open")) %>%
  mutate(year = as.integer(format(as.Date(tourney_date), "%Y"))) %>%
  select(year, tourney_name, winner_id, loser_id, winner_ioc, loser_ioc) %>%
  pivot_longer(cols = c(winner_id, loser_id), names_to = "role", values_to = "player_id") %>%
  mutate(country = ifelse(role == "winner_id", winner_ioc, loser_ioc)) %>%
  distinct(year, tourney_name, player_id, country) %>%
  count(year, country, name = "participants")

top_countries <- participants %>%
  group_by(country) %>%
  summarize(total_participants = sum(participants), .groups = "drop") %>%
  arrange(desc(total_participants)) %>%
  slice_head(n = 6) %>%
  pull(country)

filtered_data <- participants %>%
  filter(country %in% top_countries)

ggplot(filtered_data, aes(x = year, y = participants, fill = country)) +
  geom_area(alpha = 0.7) +
  labs(
    title = "Participation Trends Over Time for Top 6 Countries",
    x = "Year",
    y = "Number of Participants",
    fill = "Country"
  ) +
  theme_minimal()
```

The countries with the most players over the years are `r top_countries[1]`, `r top_countries[2]`, and `r top_countries[3]`. We can see a noticeable dip in participation to only `r participants %>% filter(year == 2020) %>% summarize(total = sum(participants)) %>% pull(total)` players in 2020, as Wimbledon was not conducted during the COVID-19 pandemic.

## Trends of Winners Over Time

Over the years, the landscape of tennis has seen significant changes in terms of who wins and how often. This first graph shows the trends in the number of titles won by players in Grand Slam tournaments, illustrating the dominance of certain players over time. 

```{r winners, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)

tennis_data <- read.csv("data/cleaned_df_webscraping_data_added.csv") %>%
  mutate(
    tourney_date = as.Date(tourney_date),
    winner_age = as.numeric(winner_age),
    loser_age = as.numeric(loser_age),
    minutes = as.numeric(minutes)
  )

top_winners_per_year <- tennis_data %>%
  filter(tourney_name %in% c("Australian Open", "Roland Garros", "Wimbledon", "US Open")) %>%
  drop_na(winner_name) %>%
  count(year = format(tourney_date, "%Y"), winner_name, name = "wins") %>%
  group_by(year) %>%
  slice_max(wins, n = 1) %>%
  ungroup()

ggplot(top_winners_per_year, aes(x = as.integer(year), y = wins, color = winner_name)) +
  geom_line() +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Trends of ATP Winners Over Time in Grand Slams",
    x = "Year", y = "Number of Wins"
  ) +
  scale_x_continuous(breaks = unique(as.integer(top_winners_per_year$year))) +
  theme(legend.position = "right")
```

- Novak Djokovic and Rafael Nadal are the most prominent players, with a consistent high number of Grand Slam wins over the years.  
- Dominic Thiem achieved a win in 2020 but hasn’t sustained dominance compared to Djokovic and Nadal.  
- Jannik Sinner appears as a new competitor, indicating the rise of younger talent breaking into Grand Slam victories in recent years.  
